Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2023-12-19 paper.tex

\subsection{Computational algorithms}
Three-point correlation function $S_3$ can be computed either in spatial domain
or in frequency domain via the convolution theorem. The simplest algorithm below
calculates $S_3$ pointwise in spatial domain:
\begin{algorithmic}[1]
  \Procedure{$S_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \State $A'_n \gets \mathfrak{S}(A_n, x_1)$
  \State $A''_n \gets \mathfrak{S}(A_n, x_2)$
  \State $T \gets A_n \cdot A'_n \cdot A''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
In this algorithm the indicator function $I_n$ for the phase of interest $n$ is
applied to the input array $A$. Then the shift operator $\mathfrak{S}$ is used
to shift $A_n$ by vectors $x_1$ and $x_2$. Finally, the three arrays are
multiplied pointwise and normalized sum of all elements of the product is
returned.

There are two common boundary conditions when applying $\mathfrak{S}$. The first
condition is to extend $A_n$ periodically (\cref{fig:s3-periodic}) when
accessing out-of-bounds array elements. This is the same as circular-shifting
$A_n$. In this case $Norm(A, x_1, x_2)$ does not depend on $x_1$ and $x_2$ and
is equal to the total number of elements in $A$. The second condition is to
replace out-of-bound elements with zeros (so-called zero padding)
(\cref{fig:s3-zeros}). A formula for $Norm(A, x_1, x_2)$ in this case is in
\ref{sec:number-of-trials}.
\begin{figure*}[tp]
  \centering
  \subfigure[Periodic boundary conditions]{
    \includegraphics[width=0.45\linewidth]{images/periodic.png}
    \label{fig:s3-periodic}}
  \hfill
  \subfigure[Zero padding]{
    \includegraphics[width=0.45\linewidth]{images/zeros.png}
    \label{fig:s3-zeros}}
  \caption[]{Computation of unnormalized $S_3$ function at point $(1, 2)$ for
    one-dimensional sequence of length 8.}
  \label{fig:s3-computation}
\end{figure*}

@@

To compute $C_3$ function we need to use a connected-component labeling
algorithm \cite{4728561,PhysRevB.14.3438} to label clusters. Then, again, a
slightly modified algorithm for $S_3$ is used to assess cluster function:
\begin{algorithmic}[1]
  \Procedure{$C_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \State $C_n \gets \mathfrak{C}(A_n)$
  \State $C'_n \gets \mathfrak{S}(C_n, x_1)$
  \State $C''_n \gets \mathfrak{S}(C_n, x_2)$
  \State $T \gets C_n \odot C'_n \odot C''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
Here $\mathfrak{C}$ is connected-component labeling operator and $\odot$ is
defined as follows:
\begin{equation}
  x \odot y = \left\{
  \begin{array}{ll}
    1 & \quad x = y \\
    0 & \quad \text{otherwise}.
  \end{array}
  \right.
\end{equation}

@@

Heatmaps of $S_3$ for the two highly anisotropic samples can be seen on
\cref{fig:heatmaps}. Anisotropicity of the both samples can be seen as a slower
decay of $S_3$ in one of the directions. Additionally, it can be observed that
elongated clusters of the void phase in the sample of soil are parallel to each
other and have a regular spacing of about $50 \mu m$.

@@

Our algorithm fixes $\theta = \frac{pi}{2}$ and does not perform random
rotations or trial triangles. It is sensitive to anisotropicity of the medium
when it is stretched along one of the axes as shown in \cref{fig:aniso}. The
main advantage of our algorithm is its parallelizability and ease of
implementation. An equivalent operation for rotation of test triangles is a
rotation of the whole input image which we have implemented in our
CorrelationFunctions.jl package. By applying a rotation to the input first we
made a rest of the algorithm cache-friendly by avoiding random memory access.
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.4\linewidth, frame]{images/aniso.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{images/aniso-s3.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{images/aniso-s3-avg.png}
  \caption[]{$S_3$ function for anisotropic media (on the left) elongated along
    the vector $(1, 1)$. Anisotropicity of the medium can be seen on a heatmap
    of $S_3$ function (in the middle). An average of 120 $S_3$ evaluations of
    the medium rotated by a random angle is on the right.}
  \label{fig:aniso}
\end{figure}

>>>>

\subsection{Computational algorithms}
Three-point correlation function $S_3$ can be computed either in spatial domain
or in frequency domain via the convolution theorem. Suppose that a finite
countable set $A$ is defined as a multidimensional array. A point $x$ belongs to
a subset $A_n$ if and only if $A[x] = n$. The simplest algorithm below
calculates $S_3$ pointwise in spatial domain:
\begin{algorithmic}[1]
  \Procedure{$S_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \Comment Apply $I_n$ to the input array elementwise.
  \State $A'_n \gets \mathfrak{S}_{x_1}(A_n)$
  \State $A''_n \gets \mathfrak{S}_{x_2}(A_n)$
  \State $T \gets A_n \cdot A'_n \cdot A''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
In this algorithm the indicator function $I_n$ for the phase of interest $n$ is
applied to the input array $A$. Then the shift operator $\mathfrak{S}$ is used
to shift $A_n$ by vectors $x_1$ and $x_2$. Finally, the three arrays are
multiplied pointwise and a normalized sum of all elements of the product is
returned.

There are two common boundary conditions when applying $\mathfrak{S}$. The
first condition is to extend $A_n$ periodically (\cref{fig:s3-periodic}) when
accessing out-of-bounds array elements. In this case $\mathfrak{S}_z$ is a
circular shift operator:
\begin{equation}
  A[x] \leftarrow A[x+z \mod d]
\end{equation}
In this case $Norm(A, x_1, x_2)$ does not depend on $x_1$ and $x_2$ and
is equal to the total number of elements in $A$. The second condition is to
replace out-of-bound elements with zeros (so-called zero padding)
(\cref{fig:s3-zeros}). A formula for $Norm(A, x_1, x_2)$ in this case is in
\ref{sec:number-of-trials} and $\mathfrak{S}_z$ is defined as follows:
\begin{equation}
  A[x] \leftarrow \left\{
  \begin{array}{ll}
    A[x+z] & \quad \text{if $x+z$ is a correct index in $A$} \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}
\begin{figure*}[tp]
  \centering
  \subfigure[Periodic boundary conditions]{
    \includegraphics[width=0.45\linewidth]{images/periodic.png}
    \label{fig:s3-periodic}}
  \hfill
  \subfigure[Zero padding]{
    \includegraphics[width=0.45\linewidth]{images/zeros.png}
    \label{fig:s3-zeros}}
  \caption[]{Computation of unnormalized $S_3$ function at point $(1, 2)$ for
    one-dimensional sequence of length 8.}
  \label{fig:s3-computation}
\end{figure*}

@@

To compute $C_3$ function we need to use a connected-component labeling
algorithm \cite{4728561,PhysRevB.14.3438} to label clusters. Then, again, a
slightly modified algorithm for $S_3$ is used to assess cluster function:
\begin{algorithmic}[1]
  \Procedure{$C_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \State $C_n \gets \mathfrak{C}(A_n)$
  \State $C'_n \gets \mathfrak{S}_{x_1}(C_n)$
  \State $C''_n \gets \mathfrak{S}_{x_2}(C_n)$
  \State $T \gets C_n \odot C'_n \odot C''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
Here $\mathfrak{C}$ is connected-component labeling operator and $\odot$ is
defined as follows:
\begin{equation}
  x \odot y = \left\{
  \begin{array}{ll}
    1 & \quad x = y \\
    0 & \quad \text{otherwise}.
  \end{array}
  \right.
\end{equation}

@@

Heatmaps of $S_3$ for the two highly anisotropic samples can be seen on
\cref{fig:heatmaps}. Anisotropy of the both samples can be seen as a slower
decay of $S_3$ in one of the directions. Additionally, it can be observed that
elongated clusters of the void phase in the sample of soil are parallel to each
other and have a regular spacing of about $50 \mu m$.

@@

Our algorithm fixes $\theta = \frac{pi}{2}$ and does not perform random
rotations or trial triangles. It is sensitive to anisotropy of the medium
when it is stretched along one of the axes as shown in \cref{fig:aniso}. A
direction of elongation for a $N$-dimensional image can be easily detected by
firstly evaluating a $N \times N$ covariance matrix for points belonging to the
phase of interest:
\begin{equation}
  C_{ij} = \sum_{x \in \left\{ \text{Coordinates of points in the
      phase of interest} \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and then decomposing $C$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation and
can be used to align the elongation with the axes (compare \cref{fig:aniso-unrot}
and \cref{fig:aniso-rot}).

The main advantage of our algorithm is its parallelizability and ease of
implementation. An equivalent operation for rotation of test triangles is a
rotation of the whole input image which we have implemented in our
CorrelationFunctions.jl package. By applying a rotation to the input first we
made a rest of the algorithm cache-friendly by avoiding random memory access.
\begin{figure}[tp]
  \centering

  \subfigure[An original image of an anisotropic pore.]{
    \includegraphics[width=0.45\linewidth, frame]{images/aniso.png}
    \label{fig:aniso-unrot}}
  \hfill
  \subfigure[A rotated image of pore, so it is elongated along the axis $X$.]{
    \includegraphics[width=0.45\linewidth, frame]{images/aniso-rot.png}
    \label{fig:aniso-rot}}
  \vskip\baselineskip
  \subfigure[$S_3$ function for the original image]{
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png}}
  \hfill
  \subfigure[$S_3$ function for the rotated image. A vector $(0.92, -0.4)$ is a
    direction of elongation of the pore.]{
  \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png}}
  \hfill
  \subfigure[$S_3$ function averaged across 100 random rotations.]{
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png}
    \label{fig:aniso-avg}}
  \caption[]{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. A result on \cref{fig:aniso-avg} is similar
    to the algorithm proposed by Malmir~et~al.}
  \label{fig:aniso}
\end{figure}
